Standardization of ceramic shape: A case studay from the Iron Age pottery from northeastern Taiwan

### 2020-08-07 analysis/paper/paper.Rmd
edit discussion

Whether standardization reflects an emergence of specialized production is associated with the number of producers, which distinguishes mechanical standardization from intentional standardization expressed by stylish and functional attributes [@Costin2001; @Costin1995]. In other words, mechanical standardization depends on potter's skills and experiences that increased morphological uniformity normally reflects fewer producers. The population could be an indicator to estimate potential number of producer. According to the historical records [@Nakamura1938, pp. 12], the population in Kiwulan reached higher number in the 17th century as the European presence and then decreased in the Chinese period due to a movement of indigenous people to the south [@Chen2007]. The decline in population is supported by ceramic abundance in total, which indicates less population might be a factor leading to higher standardization in the Chinese period.

To determine if shape changes might be related to changes in the function of pots at Kiwulan, we used geochemical methods to extract and identify lipids trapped in the fabric of potsherds to identify foods that may have contributed residues absorbed into the clay [cf. @Kwak2015]. Unfortunately, we did not obtain useful results due to extremely low lipid yields, which is probably due to the very thin, dense, and low porosity fabric of Kiwulan pottery. These physical characteristics of the clay offer limited spaces to trap and protect organic molecules from microbiological degradation [cf. @Evershed2008, pp. 909].

The spatial analysis (Figure \@ref(fig:kernel-plot)) for pottery samples presents multiple clusters with high densities of pottery during European presence. Hypothesis testing on spatial randomness indicates a non-randomly dispersed distribution before European contact and more extreme dispersed distribution after European presence. In contrast, the distribution of pottery is more similar to random distributions during the Chinese period. This contradicts our expectation that clustered pattern will be observed with an increase in pottery standardization since the emergence of specialized groups. The absence of clusters in the Chinese period is notable because this was a time of a historically-documented decline of Indigenous population [@Chen2007; @Hsieh2009]. We might expect reduced numbers of potters to result in pottery production shrinking to a few locations in the settlement during this time. Although the overall number of vessels is smaller during the Chinese contact period, Figure \@ref(fig:kernel-plot) shows that pottery is distributed randomly across the sampling area without any distinctive clusters during this time. This suggests that differences in population across our three occupation phases are probably not driving variation in craft specialization.

>>>>

Whether standardization suggests an emergence of specialized production depends on the number of producers, which distinguishes mechanical standardization from intentional standardization reflected by stylish and functional attributes [@Costin2001; @Costin1995]. In other words, mechanical standardization is defined by potter's skills and experiences that increased morphological uniformity normally relates to fewer producers. Population could be an indicator to estimate the potential number of producers. According to the historical records [@Nakamura1938, pp. 12], the population in Kiwulan reached a higher number in the 17th century as the European presence and decreased in the Chinese period due to movement of indigenous people to the south [@Chen2007]. The population dynamics in historical documents corresponds to an increase in ceramic abundance in the European presence and a decrease in the Chinese period. Thus, we model cv values as a function of the mass of ceramics using Poisson GLM with a link function. The result suggests that ceramic abundance strongly predicts the cv values (β = $`r beta_print`$, p = $`r pval_print`$). This indicates that standardized ceramics in the Chinese period may be influenced by a small population leading to fewer producers if ceramic abundance can reflect the population at Kiwulan.

In order to ensure a mechanical standardization caused by fewer producers or specialized organization, we need to investigate other factors that may contribute to intentional standardization, such as the function of ceramics. To determine if shape changes might be related to changes in the function of pots at Kiwulan, we used geochemical methods to extract and identify lipids trapped in the fabric of potsherds to identify foods that may have contributed residues absorbed into the clay [cf. @Kwak2015]. Unfortunately, we did not obtain useful results due to extremely low lipid yields, which is probably due to the very thin, dense, and low porosity fabric of Kiwulan pottery. These physical characteristics of the clay offer limited spaces to trap and protect organic molecules from microbiological degradation [cf. @Evershed2008, pp. 909].

The spatial analysis (Figure \@ref(fig:kernel-plot)) can also provide information about the intensity and location of manufacturing. The result reveals that pottery samples presents multiple clusters with high densities of pottery during European presence. Hypothesis testing on spatial randomness indicates a non-randomly dispersed distribution before European contact and more extreme dispersed distribution after European presence. In contrast, the distribution of pottery is more similar to random distributions during the Chinese period. This contradicts our expectation that clustered pattern will be observed with an increase in pottery standardization since the emergence of specialized groups. The absence of clusters in the Chinese period is notable because this was a time of a historically-documented decline of Indigenous population [@Chen2007; @Hsieh2009]. We might expect reduced numbers of potters to result in pottery production shrinking to a few locations in the settlement during this time. Although the overall number of vessels is smaller during the Chinese contact period, Figure \@ref(fig:kernel-plot) shows that pottery is distributed randomly across the sampling area without any distinctive clusters during this time. This suggests that differences in population across our three occupation phases are probably not driving variation in craft specialization.
